---
title: "DATA 505: Lab 1"
subtitle: "Getting Started with R, RStudio and Quarto"
date: "2025-08-26"
author: "" # YOUR NAME HERE (in quotes at left)
format: 
  pdf:
    toc: true
    toc-depth: 1
    number-sections: true
    output-file: lab_1.pdf
---

## Setup & Goals

Welcome to Lab 1! In this session you will:

- Learn to navigate RStudio
- Write, save, and run an R script
- See how Quarto works (our tool for mixing text and code)

> **Reminder:** Today is about getting comfortable with the environment.  
> Don’t worry if the code examples feel very simple.

If you have not installed R, RStudio, and Quarto already, see the instructions here: [Installing required software](https://andrew-bean.github.io/data505/lectures/week_1/lectures_1.html#installing-required-software).

## Navigating RStudio

When you open RStudio, you’ll see **four panes**. The official [Getting Started guide](https://docs.posit.co/ide/user/ide/get-started/#rstudio-panes) describes their purposes:

1. The Source pane is where you can edit and save R or Python scripts or author computational documents like Quarto and R Markdown.

2. The Console pane is used to write short interactive R commands.

3. The Environment pane displays temporary R objects as created during that R session.

4. The Output pane displays the plots, tables, or HTML outputs of executed code along with files saved to disk.

![From [Getting Started guide](https://docs.posit.co/ide/user/ide/get-started/#rstudio-panes)](figures/rstudio_layout.jpeg)

### Exercise

- Find the **Console** (bottom-left).  
- Type:

```r
2 + 2
```

## Writing scripts and running code

The Console is for quick experiments. For work you want to save and rerun, you should use a script.

### Demo

1. Create a new script:
    - `File -> New File -> R Script`
    - Using the editor window, type the code in @mean-numbers
    - Save the file as `data505_lab1.R`
    - You have created your first R source file!

```{r}
#| label: mean-numbers
#| echo: true
#| eval: false
#| number-lines: true

c(2, 4, 6, 8, 10)
numbers <- 2 * (1:5)
print(numbers)
mean(numbers)
sample_mean <- mean(numbers)
```


2. Now you will practice executing code in RStudio. A useful reference: ["Executing Code"](https://docs.posit.co/ide/user/ide/guide/code/execution.html) in the RStudio User Guide
    - Run the first line of @mean-numbers. What happens in the console?
    - Now run the next three lines one at a time. What variables are now present in the environment?
    - Next restart your R session (`Session -> Restart R`, or one of [these keyboard shortcuts](https://docs.posit.co/ide/user/ide/reference/shortcuts.html#session)). What happened to the variables in the environment?
    - Now practice running a selection of code (multiple selected lines). Select the lines and then do `Ctrl + Enter` / `Cmd + Enter` (or click "Run" from the toolbar).
    - Finally, run the entire file using the "Source" button or `Ctrl + Shift + Enter`

## Quarto demo

We will be using Quarto for lab assignments and case studies. Quarto lets us mix text (like these instructions) with code and output within a single script (a `.qmd` file). 

Code is written in **code chunks**, which are executed when the document is **rendered**. The result is a formatted document which includes the code, the output, along with all the surrounding text.

We will get more practice with Quarto next week.

### Exercise

- Download the Quarto script for Lab 1 from here: [lab_1_blank.qmd](https://andrew-bean.github.io/data505/labs/week_1/lab_1.qmd)
- Open it in RStudio
- Edit the script to add yourself as author
- Edit the code chunk below to print your name (use `print()`)
- Render the script 
- Submit your edited code AND the pdf output to Moodle: [Link](https://moodle.drew.edu/mod/assign/view.php?id=594553)

```{r}

```
